Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-06-05 main.tex
Intro improvements and fixed some units.

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large number of
reported injuries~\citep{Wegman2024}. At low speeds, from start up to typical
cruising speeds, bicycles themselves are not self-stable and can be challenging
for the rider to balance. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds by  relieving the rider of a necessary control
activity. Bicycles can be mechanically modified to lower the speeds at which
they are self-stable~\citep{Astrom2005} and since the 1980s it has been known
that applying a motor actuated steering torque proportional to the roll rate
can stabilize a single track vehicle at very low speeds. If this automatic
control of steering can stabilize a bicycle, it may reduce the control required
by the rider to successfully manage balancing tasks. We have developed such a
balance assisting bicycle and hypothesize that it helps the rider in situations
they are likely to fall, Figure~\ref{fig:balance-assist-bicycle}.
%
\begin{figure}
  \centering
  \includegraphics[width=70mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

In the early years of developments in automatic control, \citep{Whipple1899}
remarkably realized and showed that roll motion feedback can stabilize a
bicycle. Much later, automatic roll stabilization of single track vehicles
began in earnest after predictive motorcycle models were developed and refined
in throughout the 1970s. \Citet{vanZytveld1975} was influenced by Whipple's
work and seems to be the first to attempt to robotically stabilize a small
motorbike with a controlled inverted pendulum that mimicked rider lean, but he
was not successful in demonstrating what his theoretic control model correctly
predicted. In his model, he recognized that feedback of the vehicle roll angle
and angular rate was essential to stabilize the vehicle. It was not until the
early 1980s that \cite{Nagai1983} successfully demonstrated balancing a robotic
bicycle on a treadmill with both steering control and a laterally moving mass.
Ruijs and Pacjecka followed this by demonstrating an automatically balanced
motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
proposed how the gains must change with respect to vehicle speed for favorable
control at all speeds. This roll motion feedback is the simplest controller
that can stabilize a single track vehicle above a minimum speed when one is not
concerned with wobble instabilities, something not prominent in bicycles at low
speeds. Ruijs and Pacejka's work was not particularly concerned with low speed
stability and their vehicle was fully automatic, no human rider was involved.
Many more automatically balanced single track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in balancing and possibly reduce single actor falls. Most
of these robotic bicycle and motorcycle designers did not intend for a human
rider to also control the stabilized vehicle. Nevertheless, an automatically
stabilized bicycle can be controlled by a human rider if the motor controlled
steer torque and the rider applied steer torque act on the steer in parallel.
The effect of this automatic control gives the ability to change the dynamics,
up to some limits, of the human controlled plant. In our prior
study~\citep{Alizadehsaravi2023}, we demonstrate reduced motion during
distractions due to the balance assist but contrarily researchers with a
similar vehicle recently showed rider dissatisfaction with the
stabilization~\cite{Hanakam2023}.

The linear Carvallo-Whipple bicycle model~\citep{Carvallo1899,Whipple1899} is
the simplest bicycle model that exhibits non-minimum phase behavior and
self-stability. The bicycle model is ideal for showing the effect of a roll
motion feedback driven steer motor on the dynamics. The linear model can be
described by the fourth order state space equations:

\begin{align}
  \dot{\vec{x}} = \mathbf{A} \vec{x} +
  \mathbf{B}
  \vec{u}
  \textrm{ where }
  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
  \textrm{ and }
  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
  \textrm{.}
\end{align}
%
The states are the roll angle \(\phi\) and steer angle \(\delta\) along with
their derivatives and the inputs are roll torque \(T_\phi\) and steer torque
\(T_\delta\). The state \(\mathbf{A}\) and input \(\mathbf{B}\) matrices are
functions of the equilibrium forward speed \(v\) and  are otherwise populated
with expressions that are functions of the geometric and inertial parameters of
the nonholonomic multibody system made up of four rigid bodies: two wheels,
front frame, and rear frame.

@@

The state matrix \(\mathbf{A}\) and input matrix \(\mathbf{B}\) are both
functions of the equilibrium speed \(v\) and \(k_{\dot{\phi}}\) can be selected
such that the eigenvalues of \(\left(\mathbf{A} - \vec{B}_\delta\left[0 \quad
k_{\dot{\phi}} \quad 0 \quad 0\right] \right)\) have negative real parts for
\(v_{min} < v < v_{capsize}\). With gain scheduling with respect to \(v\), the
speed range where the bicycle is stable can be maximized within any physical
actuator magnitude and bandwidth limits. \citet{Schwab2008} elaborates on some
of the possibilities in scheduling the gains for such a controller and shows
that a linear scheduling with respect to speed can give satisfactory stability.

In this paper, we test whether a gain scheduled steer motor controlled bicycle
that is stable in a large low speed range (when riderless) assists the rider in
balancing by investigating whether it is beneficial in preventing falling.  To
do this, we apply varying magnitude mechanical perturbations to the handlebars
while the rider is balancing the bicycle on a treadmill and then assess the
rider's probability of falling with the balance assist steer motor system on
and off.

\todo{Talk about the types of real life perturbations an how our perturbation
mimics it.}

\section{Methods}
%
\subsection{Bicycle}
%
We modified a standard electric bicycle (Gazelle Grenoble C7\todo{Double check
this is the right model}, Dieren, The Netherlands) with a custom motor in the
steering assembly capable of applying up to 7~\si{\newton\meter} of torque
between the head tube and steer tube, Figure~\ref{fig:balance-assist-bicycle}.
A custom motor controller converts the commanded torque to applied torque. We
measure the speed of the rear wheel with an encoder (Magura XXX) and measure
the body fixed roll rate of the bicycle with a MEMs rate gyroscope (Company,
Location). The balance assist control algorithm is implemented on a
microprocessor (Teensy, Company, Location) and data from all sensors is logged
with a CAN bus (Company, Location) at approximately X Hz.

\subsection{Balance Assist Control}
We use a forward speed \(v\) gain scheduled proportional roll rate feedback
controller to stabilize the bicycle. The commanded steer torque
\(T^\textrm{m}_\delta\) from the steer motor follows the control law
\begin{align}
  T^\textrm{m}_\delta = k(v_{weave} - v)\dot{\phi}
\end{align}
where 4.7~\si{\meter\per\second} is the weave critical speed predicted from the
open loop bicycle rigid-rider dynamics. We use \(k=8\) and \(k=10\) as gain
values during the experiments. Scaling the proportional feedback gain linearly
with respect to speed stabilizes the normally unstable weave mode of the
bicycle from speeds of X~\si{\kilo\meter\per\hour} to
X~\si{\kilo\meter\per\hour} as shown in Figure~\ref{fig:eigenvalues}.

@@

\subsection{Subjects}
%
We recruited 26 able-bodied young adults (20-36 years old) to participate in
the experiments. The subjects all were confident in their cycling skills and
had cycled at least once in the last month. Eleven subjects performed the
experiments at 1.7~\si{\meter\per\second} (6~\si{\kilo\meter\per\hour}) and
fifteen subjects performed the experiments at 2.8~\si{\meter\per\second}
(10~\si{\kilo\meter\per\hour}). All subjects consented to the experiment and
could decline to continue at any time. The study was approved by Delft
University of Technology's Human Research Ethics Council (\#XXX).

@@

\subsection{Protocol}
%
The subjects were divided into two groups. The first group performed the
protocol at a belt speed of 2.8~\si{\meter\per\second}
(10~\si{\kilo\meter\per\hour}) with the controller gain at $k=8$ and the second
group performed the protocol at a belt speed of 1.7~\si{\meter\per\second}
(6~\si{\kilo\meter\per\hour}). The 10~\si{\kph} group's experiments occurred X
weeks before the 6~\si{\kilo\meter\per\hour} group.

@@

Five possible perturbation forces are determined based on the initial
perturbation threshold estimate: the initial estimate itself, two perturbations
lower than the initial estimate and two perturbations higher than the initial
estimate.  The five possible perturbations are separated by 10 N steps. For
example, if the initial estimate of the perturbation threshold of a participant
is 80 N, the five possible perturbations are 60, 70, 80, 90 and 100 N. Which
one of these five perturbations is chosen is determined at random. If the
perturbation results in a fall, the estimate of the perturbation threshold is
decreased with 10 N, and vice versa if the perturbation did not result in a
fall. Five new possible perturbation forces are determined around the updated
perturbation threshold, and a new perturbation is selected at random. This
process iterates until twenty perturbations are applied.

@@

At the initiation of each perturbation we log the steer and roll angles to
characterize the configuration of the bicycle when perturbed. The gain setting
on the balance assist controller indicates if the assistance is off \(k=0\) or
on at two different levels: low \(k=8\) or high \(k=10\). A recovery from the
perturbation is successful if the person neither put their foot down onto the
treadmill surface or either wheel of the bicycle exits the width of the
treadmill belt. We record this as a binary variable \(f\). All measured
variables are reported in Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall outcome'' and
    ``Perturbation order number'' are recorded per perturbation instance.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain & \(k\) & TODO & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
    Fall outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear & \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) &  \si{\degree} & Kalman estimate~\cite{Gabriel2023} \\
    Roll Angular Rate & \(\dot{\phi}\) &  \si{\degree\per\second} & rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

@@

We evaluate this hypothesis using a multivariate logistic regression model that
takes the general form
%
\begin{align}
  f_{ij} | p_{ij} \sim \textrm{Bern}(p_{ij}) \textrm{.}
\end{align}
%
Fall outcome \(f_{ij}\) is the binary outcome of perturbation \(j\) on
participant \(i\) which follows a Bernoulli distribution given the probability
\(p_{ij}\) that a fall occurs. The log-odds of the probability is then a linear
function of our independent variables with \(\beta\) as the intercept and
\(\alpha_k\) as the linear coefficients to the independent variables
\(x_{ij}^k\), i.e. all variables in Table~\ref{tab:statmodel-variables} except
\(f\).
%
\begin{align}
  \log \left(\frac{p_{ij}}{1-p_{ij}} \right) = \beta + \sum_{k=0}^{K} \alpha_k x_{ij}^{k}
\end{align}

>>>>

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large number of
reported injuries~\citep{Wegman2024}. At low speeds, from start up to typical
cruising speeds, bicycles themselves are not self-stable and can be challenging
for the rider to balance. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds by  relieving the rider of a necessary control
activity. Bicycles can be mechanically modified to lower the speeds at which
they are self-stable~\citep{Astrom2005} and since the 1980s it has been known
that applying a motor actuated steering torque proportional to the roll rate
can stabilize a single track vehicle at very low speeds. If this automatic
control of steering can stabilize a bicycle, it may reduce the control required
by the rider to successfully manage balancing tasks. We have developed such a
balance assisting bicycle and hypothesize that it helps the rider in situations
they are likely to fall, Figure~\ref{fig:balance-assist-bicycle}.
\todo{Talk about the types of real life perturbations an how our perturbation
mimics it.}
%
\begin{figure}
  \centering
  \includegraphics[width=70mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

In the early years of developments in automatic control, \citep{Whipple1899}
not only derived the correct equations of motion of the bicycle but realized
and showed that roll motion feedback can stabilize a bicycle. Much later,
automatic roll stabilization of single track vehicles began in earnest after
predictive motorcycle models were developed and refined in throughout the
1970s. \Citet{vanZytveld1975} was influenced by Whipple's work and seems to be
the first to attempt to robotically stabilize a small motorbike with a
controlled inverted pendulum that mimicked rider lean, but he was not
successful in demonstrating what his theoretic control model correctly
predicted. In his model, he recognized that feedback of the vehicle roll angle
and angular rate was essential to stabilize the vehicle. It was not until the
early 1980s that \cite{Nagai1983} successfully demonstrated balancing a robotic
bicycle on a treadmill with both steering control and a laterally moving mass.
Ruijs and Pacjecka followed this by demonstrating an automatically balanced
motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
proposed how the gains must change with respect to vehicle speed for favorable
control at all speeds. This roll motion feedback is the simplest controller
that can stabilize a single track vehicle above a minimum speed when one is not
concerned with wobble instabilities, something not prominent in bicycles at low
speeds. Ruijs and Pacejka's work was not particularly concerned with low speed
stability and their vehicle was fully automatic, no human rider was involved.
Many more automatically balanced single track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in balancing and possibly reduce single actor falls. Most
of these robotic bicycle and motorcycle designers did not intend for a human
rider to also control the stabilized vehicle. Nevertheless, an automatically
stabilized bicycle can be controlled by a human rider if the motor controlled
steer torque and the rider applied steer torque act on the steer in parallel.
The effect of this automatic control gives the ability to change the dynamics,
up to some limits, of the human controlled plant. In our prior
study~\citep{Alizadehsaravi2023}, we demonstrate reduced motion during
distractions due to the balance assist but contrarily researchers with a
similar vehicle recently showed rider dissatisfaction with the
stabilization~\cite{Hanakam2023}.

The linear Carvallo-Whipple bicycle model~\citep{Carvallo1899,Whipple1899} is
the simplest bicycle model that exhibits non-minimum phase behavior and
self-stability. The bicycle model is ideal for showing the effect of a roll
motion feedback driven steer motor on the dynamics. The linear model can be
described by the fourth order state space equations:
%
\begin{align}
  \dot{\vec{x}} = \mathbf{A} \vec{x} +
  \mathbf{B}
  \vec{u}
  \textrm{ where }
  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
  \textrm{ and }
  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
  \textrm{.}
\end{align}
%
The states are the roll angle \(\phi\) and steer angle \(\delta\) along with
their derivatives and the inputs are roll torque \(T_\phi\) and steer torque
\(T_\delta\). The state \(\mathbf{A}\) and input \(\mathbf{B}\) matrices are
functions of the equilibrium forward speed \(v\) and  are otherwise populated
with expressions that are functions of the geometric and inertial parameters of
the nonholonomic multibody system made up of four rigid bodies: two wheels,
front frame, and rear frame.

@@

The state matrix \(\mathbf{A}\) and input matrix \(\mathbf{B}\) are both
functions of the equilibrium speed \(v\) and \(k_{\dot{\phi}}\) can be selected
such that the eigenvalues of \(\left(\mathbf{A} - \vec{B}_\delta\left[0 \quad
k_{\dot{\phi}} \quad 0 \quad 0\right] \right)\) have negative real parts for
\(v_{min} < v < v_\textrm{capsize}\) where \(v_{min}\) is the lowest stable
speed given \(k_{\dot{\phi}}\) and \(v_\textrm{capsize}\) is the speed at which
the uncontrolled bicycle's capsize mode goes unstable. With gain scheduling
with respect to \(v\), the speed range where the bicycle is stable can be
maximized within any physical actuator magnitude and bandwidth limits.
\citet{Schwab2008} elaborates on some of the possibilities in scheduling the
gains for such a controller and shows that a linear scheduling with respect to
speed can give satisfactory stability.

In this paper, we test whether a linearly gain scheduled steer motor controlled
bicycle that is stable in a large low speed range (when riderless) assists the
rider in balancing by investigating whether it is beneficial in preventing
falling.  To do this, we apply varying magnitude mechanical perturbations to
the handlebars while the rider is balancing the bicycle on a treadmill and then
assess the rider's probability of falling with the balance assist steer motor
system on and off.

\section{Methods}
%
\subsection{Bicycle}
%
We modified a standard electric bicycle (Gazelle Grenoble C7\todo{Double check
this is the right model}, Dieren, The Netherlands) with a custom motor in the
steering assembly capable of applying up to 7~\si{\newton\meter} of torque
between the head tube and steer tube, see
Figure~\ref{fig:balance-assist-bicycle}.  A custom motor controller converts
the commanded torque to applied torque. We measure the speed of the rear wheel
with an encoder (Magura XXX) and measure the body fixed roll rate of the
bicycle with a MEMs rate gyroscope (Company, Location). The balance assist
control algorithm is implemented on a microprocessor (Teensy, Company,
Location) and data from all sensors is logged with a CAN bus (Company,
Location) at approximately X Hz.

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll rate feedback
controller to stabilize the bicycle. The commanded steer torque
\(T^\textrm{m}_\delta\) from the steer motor follows the control law
%
\begin{align}
  T^\textrm{m}_\delta = k(v_{weave} - v)\dot{\phi}
\end{align}
%
where 4.7~\si{\meter\per\second} is the weave critical speed predicted from the
open loop bicycle rigid-rider dynamics. We use \(k=8\) and \(k=10\) as gain
values during the experiments. Scaling the proportional feedback gain linearly
with respect to speed stabilizes the normally unstable weave mode of the
bicycle from speeds of X~\si{\kilo\meter\per\hour} to
X~\si{\kilo\meter\per\hour} as shown in Figure~\ref{fig:eigenvalues}.

@@

\subsection{Protocol}
%
We recruited 26 able-bodied young adults (20-36 years old) to participate in
the experiments. The subjects all were confident in their cycling skills and
had cycled at least once in the last month. Eleven subjects performed the
experiments at 1.7~\si{\meter\per\second} (6~\si{\kilo\meter\per\hour}) and
fifteen subjects performed the experiments at 2.8~\si{\meter\per\second}
(10~\si{\kilo\meter\per\hour}). All subjects consented to the experiment and
could decline to continue at any time. The study was approved by Delft
University of Technology's Human Research Ethics Council (\#XXX).

@@

The subjects were divided into two groups. The first group performed the
protocol at a belt speed of 2.8~\si{\meter\per\second}
(10~\si{\kilo\meter\per\hour}) with the controller gain at $k=8$ and the second
group performed the protocol at a belt speed of 1.7~\si{\meter\per\second}
(6~\si{\kilo\meter\per\hour}). The 10~\si{\kph} group's experiments occurred X
weeks before the 6~\si{\kilo\meter\per\hour} group.

@@

Five possible perturbation forces are determined based on the initial
perturbation threshold estimate: the initial estimate itself, two perturbations
lower than the initial estimate and two perturbations higher than the initial
estimate.  The five possible perturbations are separated by 10 N steps. For
example, if the initial estimate of the perturbation threshold of a participant
is 80~\si{\newton}, the five possible perturbations are 60, 70, 80, 90 and
100~\si{\newton}. Which one of these five perturbations is chosen is
determined at random. If the perturbation results in a fall, the estimate of
the perturbation threshold is decreased with 10~\si{\newton}, and vice versa
if the perturbation did not result in a fall. Five new possible perturbation
forces are determined around the updated perturbation threshold, and a new
perturbation is selected at random. This process iterates until twenty
perturbations are applied.

@@

At the initiation of each perturbation we log the steer and roll angles to
characterize the configuration of the bicycle when perturbed. The gain setting
on the balance assist controller indicates if the assistance is off \(k=0\) or
on at two different levels: low \(k=8\) or high \(k=10\). A recovery from the
perturbation is successful if the person neither put their foot down onto the
treadmill surface or either wheel of the bicycle exits the width of the
treadmill belt. We record this as a binary variable \(f\). All measured
variables are reported in Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall outcome'' and
    ``Perturbation order number'' are recorded per perturbation instance.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain & \(k\) & TODO & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
    Fall outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear & \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) &  \si{\degree} & Kalman estimate~\citep{Gabriel2023} \\
    Roll Angular Rate & \(\dot{\phi}\) &  \si{\degree\per\second} & rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

@@

We evaluate this hypothesis using a multivariate logistic regression model that
takes the general form
%
\begin{align}
  f_{ij} | p_{ij} \sim \textrm{Bern}(p_{ij}) \textrm{.}
\end{align}
%
Fall outcome \(f_{ij}\) is the binary outcome of perturbation \(j\) on
participant \(i\) which follows a Bernoulli distribution given the probability
\(p_{ij}\) that a fall occurs. The log-odds of the probability is then a linear
function of our independent variables with \(\beta\) as the intercept and
\(\alpha_k\) as the linear coefficients to the independent variables
\(x_{ij}^k\), i.e. all variables in Table~\ref{tab:stat-model-variables} except
\(f\).
%
\begin{align}
  \log \left(\frac{p_{ij}}{1-p_{ij}} \right) = \beta + \sum_{k=0}^{K} \alpha_k x_{ij}^{k}
\end{align}
